Discovering causal factors with back-to-back regression

### 2019-05-22 paper/meg_exps.tex
exps

\subsection{Dataset}

Next, we apply our method to brain imaging data from the anonymized multimodal
neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset \cite{schoffelen2019204}.
The dataset contains magneto-encephalography (MEG) recordings of 102 healthy native-Dutch
adults who participated in a reading task.

Subjects were exposed to a rapid serial visual presentation of Dutch words. The word lists
consisted of 120 sentences, and scrambled lists of the same words. Each word was presented on the computer screen for 351ms on average (min: 300ms, max: 1400ms).
Successive words were separated by a blank screen for 300ms, and successive
sentences were separated by an empty screen for a few (3-4) seconds.

@@

The raw MEG data was bandpass-filtered between 0.1 and 40 Hz using MNE-Python default
parameters \cite{mne}. Specifically, we used a zero-phase finite impulse response filter (FIR) with a hamming window and with transition bands of 0.1 Hz and 10 Hz for the low and high
cut-off frequencies respectively.

@@

For each word, each subject, and each time sample relative to word onset, we build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words by $q=301$ MEG channels (273 magnetometers and 28 compensation channels). Each of the columns of $Y$ is normalized to have zero mean and unit variance.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=1.5cm 1cm 0.5cm 0cm, clip=True]{figure_meg.pdf}
  \caption{Average magnetosensor response across words for one subject.}
  \label{fig:megavg}
\end{figure}

\subsection{Word features}
We annotated each word with 38 features: 26 corresponds to the number of each letter in the word (from 'a' to 'z', independently of their case), the total number of letters ('word length'), the frequency of the word in natural language (using the Zipf logarithmic scale \cite{marc_brysbaert}), and the part-of-speech tags (i.e. NOUN, PROPN (proper noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ (adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy \cite{spacy} and WordFreq \cite{speer}. For clarity purposes, we refer to these features using two broad categories: visual (i.e. each letter and the total number of letters) and lexical (part-of-speech and word frequency).
This procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by $d=38$ features for each subject. Each of the columns of $X$ is normalized to have a mean and a standard deviation of 0 and 1 respectively.

\subsection{Analyses}
We aimed to identify which factors of $X$ explain the MEG recordings $Y$ at each time sample and for each subject separately. To this aim, we implemented three types of analyses (forward, backward and back-to-back regressions), and tested whether their results were consistent across subjects, using a second-level statistics described below.

Forward regression, also known as 'encoding' in neuroimaging, here consisted in fitting a Ridge-regularized OLS from $X$ to $Y$:
\begin{equation}
H = (X^{T}X+\lambda I)^{-1} X^{T}Y
\end{equation}
where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmatically distributed between $10^{-5}$ and $10^5$. We aimed to estimate the contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d, q}$ is a matrix of $d$ word features by $q$ MEG channels. We thus approximated the relative contribution of each feature by sum-squarring the H coefficients:
\begin{equation}
\hat e_d = \sum_{i=0}^{i=q} {h_d^i}^2
\end{equation}

Backward regression, also known as 'decoding' in neuroimaging, here consisted in fitting a Ridge-regularized OLS from $Y$ to $X$:

\begin{equation}
G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X
\end{equation}

where $\lambda$ is the regularization parameter fit with the leave-one-out cross-validation
implemented in scikit-learn \cite{sklearn} from 20 possible value logarithmically distributed
between $10^{-5}$ and $10^5$. We aimed to estimate the contribution $diag(\hat E)
\in\mathbb{D}^{d} $ of each factor. Using 10-fold cross-validation, we thus assessed the ability
of this ridge regression to predict independent and identically distributed data:

\begin{equation}
\hat X_{test} = Y_{test}\cdot G_{train}
\end{equation}

and assessing, for each factor of X independently, the Pearson correlation coefficient $r$ between $X_{test}$ and $\hat X_{test}$ for each cross-validation split.

Finally, back-to-back regression was fit as decribed earlier using two additional stability tricks designed to minimize the impact of bagging.
First the $G$ ridge regression was broken-down with SVD blablabla
Second, $X$ was regularized using a maximal likelihood blablabla
Third, no regularization

@@

\begin{equation}
$ttest(\hat E_{t=-100} - \hat E_{\tau}$
\end{equation}
with $\tau \in $ 150ms and 400ms respectively

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=1.5cm 1cm 0.5cm 0cm,
clip=True]{meg_result.pdf}
  \caption{Average magnetosensor response across words for one subject.}
  \label{fig:megresult}
\end{figure}

@@

\subsubsection{Overview}

>>>>

\newpage
\subsection{Dataset}

Next, we apply our method to brain imaging data from the anonymized multimodal
neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset
(\cite{schoffelen2019204}).  The dataset contains magneto-encephalography (MEG)
recordings of 102 healthy native-Dutch adults who participated in a reading
task.
%
Subjects were exposed to a rapid serial visual presentation of Dutch words. The
word lists consisted of 120 sentences, and scrambled lists of the same words.
Each word was presented on the computer screen for 351ms on average (min: 300ms,
max: 1400ms).  Successive words were separated by a blank screen for 300ms, and
successive sentences were separated by an empty screen for a few (3-4) seconds.

@@

The raw MEG data was bandpass-filtered between 0.1 and 40 Hz using MNE-Python
default parameters (\cite{gramfort2013meg, gramfort2014mne}). Specifically, we used a zero-phase finite impulse
response filter (FIR) with a Hamming window and with transition bands of 0.1 Hz
and 10 Hz for the low and high cut-off frequencies respectively.

@@

For each subject and each time sample relative to word onset, we
build an observation matrix $Y \in \mathbb{R}^{n, q}$ of $n\approx$ 2700 words
by $q=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
of the columns of $Y$ is normalized to have zero mean and unit variance.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_sensors.pdf}
  \caption{Magnetosensor response (femtoteslas fT) averaged across words for the first subject. The color coding corresponds to the positions of the sensors on the head as shown on the top-left diagram. The diverging curves correspond to artifacts in data collection.}
  \label{fig:megavg}
\end{figure}

\subsection{Word features} We annotated each word with 38 features: 26
correspond to the count of each letter in the word (from 'a' to 'z',
independently of their case), the total number of letters ('word length'), the
frequency of the word in natural language (using the Zipf logarithmic scale
of \cite{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
(adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy
(\cite{spacy2}) and WordFreq (\cite{speerwordfreq}). For clarity purposes, we refer to these
features using two broad categories: visual (i.e. each letter and the total
number of letters) and lexical (part-of-speech and word frequency).  This
procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by
$d=38$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

\subsection{Analyses} We attempt to identify which factors of $X$ explain the MEG
recordings $Y$ at each time sample and for each subject separately. To this aim,
we implement three types of analyses (forward, backward and back-to-back
regressions). Corresponding results can be found in Figure \ref{fig:meg_twocurves}.

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
  \caption{We compare the encoding approach (columns 1 and 2, ridge regression $X \rightarrow Y$) and the decoding approach (column 3, ridge regression $Y \rightarrow X$) to our method (column 4, $X \rightarrow Y \rightarrow X$) on real MEG data from the MOUS dataset. Top row: the first subject. Bottom row: average across subjects.
  We plot the contributions of the \textit{word frequency} and the \textit{word
  length} over time. Our method can disambiguate them consistently with
  previous neuroscientific work showing an early response ($\approx$ 150ms) to
  length and a later response ($\approx$ 400ms) to frequency. See text for
  details.}
  \label{fig:meg_twocurves}
\end{figure}

\paragraph{Forward regression} also known as 'encoding' in neuroimaging, here consists of
fitting a ridge-regularized ordinary least squares (OLS) regression \footnote{We use scikit-learn (\cite{sklearn}).} from $X$ to $Y$:
\begin{equation}
  H = (X^{T}X+\lambda I)^{-1} X^{T}Y
\end{equation}
where $\lambda$ is the
regularization parameter, cross-validated between $10^{-5}$ and $10^5$.

We then estimate the contribution
$diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d,
q}$ is a matrix of $d$ word features by $q$ MEG channels. We approximate
the relative contribution of each feature with the sum-of-squares of the corresponding coefficients in $H$; for the $j$-th feature:
\begin{equation}
  \hat e_j = \sum_{i=0}^{q} {h_{i,j}}^2
\end{equation}
\paragraph{Backward regression} or 'decoding', is similar but predicts $X$ from $Y$:
\begin{equation} G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X \end{equation}
where $\lambda$ is cross-validated as for encoding. We estimate the
contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor in the following
way. We predict $X$ on a held-out split \textit{(train, test)}:
% Using 10-fold cross-validation, we thus assessed the ability of this ridge
% regression to predict independent and identically distributed data:
\begin{equation} \hat X_{test} = Y_{test}\cdot G_{train} \end{equation}
and assess, for each factor of X independently, the Pearson correlation
coefficient $r$ between $X_{test}$ and $\hat X_{test}$. We repeat the procedure
for 20 random splits and average the result.

% Finally,
\paragraph{Back-to-back regression} was fit as described earlier using two
additional stability tricks designed to minimize the impact of bagging.  First
the $G$ ridge regression was broken-down with SVD blablabla Second, $X$ was
regularized using a maximal likelihood blablabla Third, no regularization

@@

\begin{equation} ttest(\hat E_{t=-100} - \hat E_{\tau}) \end{equation} with
$\tau \in $ 150ms and 400ms respectively

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_result.pdf}
  \caption{Average magnetosensor response across words for one subject.}
  \label{fig:megresult}
\end{figure}

@@
